Data Analysis Methods for the Detection of the Epoch of Reionization

### 2017-11-15 PAPER.tex
-Finished Danny's smaller edits throughout paper
made some of Aaron's edits for signal loss methods/practice sections

To capture the full statistical likelihood of signal loss, one requires a quick way to generate many realizations of simulated $21$ cm signal visibilities. Here we use the same method used in \citet{ali_et_al2015}, where gaussian noise visibilities are filtered in fringe-rate space to retain only ``sky-like" time-modes. This signal is then added to the data.  \footnote{One specific change from \citet{ali_et_al2015} is that we add this simulated signal into the chain before the final fringe-rate filter is applied to the data. Previously, the addition was done after that final fringe-rate filter step.  This change results in an increased estimate of signal loss, likely due to the use of the fringe-rate filter as a simulator. Whether the loss would be estimated at the same level if different $21$ cm simulations were used remains an open question.}

@@

We are interested in quantifying how much $\textbf{e}$ is lost after weighting $\textbf{r}$ and applying OQE formalism. We investigate this by comparing two quantities we define as the input power spectrum and output power spectrum: $P_{in}$ and $P_{out}$. $P_{in}$ represents the unweighted power spectrum of only $\textbf{e}$, our simulated EoR signal. $P_{out}$ is the weighted power spectrum of $\textbf{e}$ that would result from our pipeline if the signal was mixed with our data. Comparing the two quantities yields insight into how much of $\textbf{e}$ is lost due to our choice of weighting. Ignoring normalization factors, these two quantities can be written as:

@@

\begin{eqnarray}
\label{eq:sigloss}
P_{out,\alpha} \equiv \hat{\textbf{p}}_{e,\alpha} &=& \hat{\textbf{p}}_{r,\alpha}-\hat{\textbf{p}}_{x,\alpha} \nonumber \\
&\propto& \textbf{r}^{\dagger}\textbf{R}_{r}\textbf{Q}^{\alpha}\textbf{R}_{r}\textbf{r} - \textbf{x}^{\dagger}\textbf{R}_{x}\textbf{Q}^{\alpha}\textbf{R}_{x}\textbf{x}.
\end{eqnarray}

@@

One may wonder why $P_{out}$ cannot be computed simply as the weighted power spectrum of $\textbf{e}$ alone, namely $P_{out,\alpha} \propto \textbf{e}^{\dagger}\textbf{R}_{r}\textbf{Q}^{\alpha}\textbf{R}_{r}\textbf{e}$. Expanding Equation \eqref{eq:sigloss}, we see that $P_{out}$ is comprised of multiple terms:

@@

Taking the case of very large $\textbf{e}$ so that any terms involving only $\textbf{x}$ are small, yields:

@@

We see that our naive expression for $P_{out}$ is the first term in Equation \eqref{eq:pout_expand}, but there are also two additional terms. An initial assumption would be that the cross-terms that involve both $\textbf{e}$ and $\textbf{x}$ should be zero, since foregrounds and the cosmological signal are physically unrelated. In \citet{ali_et_al2015}, this assumption is used, and the first term in Equation \eqref{eq:pout_expand} is directly compared to the input (unweighted) power spectrum to compute signal loss.

However, deeper investigations of these terms reveal that they contain non-negligible power. In fact, foreground modes and signal modes are anti-correlated on average, as derived and explained in \citet{switzer_et_al2015}. The spurious correlation of the two means that we cannot compute signal loss using a signal-only simulation, which would yield greater values for $P_{out}$ and thereby underestimate signal loss. Therefore, in our revised signal loss computation we use the full quantity for $P_{out}$ as defined in Equation \eqref{eq:sigloss}, which subtracts the weighted power spectrum of the data from the weighted power spectrum of data plus EoR.

The relationship between the input and output power spectra, $P_{in}$ and $P_{out}$, can be thought of as a transfer function which, for a sampling of $P_{in}$ and $P_{out}$ provides a mapping from an output power spectrum distribution into an input distribution. By viewing data through this signal loss lens, we are asking the question ``what input power spectrum distribution could this (signal-loss affected) data come from?" In the next section, we describe our signal loss transfer function for PAPER-64 data and detail the computation used to translate our power spectrum result into one viewed through a signal loss lens.

@@

We now shift our attention towards computing signal loss for the fringe-rate filtered PAPER-64 dataset. While our methodology outlined below is robust to any weighting scheme, here we demonstrate the computation using inverse covariance weighting ($\textbf{R} = \textbf{C}^{-1}$), the weighting scheme used in \citet{ali_et_al2015} which leads to substantial loss. With this weighting, our expressions for $P_{in}$ and $P_{out}$ become:

@@

Our signal loss transfer function for one $k$ value is shown in Figure \ref{fig:sigloss_transfercurve} and compares the weighted case ($\textbf{R} = \textbf{C}^{-1}$, left plot) with an unweighted case ($\textbf{R} = \textbf{I}$, right plot). The figures depict a smoothed-out distribution (via kernel density estimators) of our input and output power spectra. To sample the distribution, we run $40$ total injection levels, where the amplitude of \textbf{e} is increased in even increments in log-space and ranges from $P_{in} \sim -10^{13} \, mK^{2} (h^{-1} Mpc)^{3}$ (where one copy of \textbf{e} in Equation \eqref{eq:Pin} is negative and the other is positive) to $P_{in} \sim10^{13} \, mK^{2} (h^{-1} Mpc)^{3}$. For each injection, we perform $20$ bootstraps (over baselines, as explained in Section \ref{sec:Boot}), resulting in a total of $800$ data points on our $P_{in}$ vs. $P_{out}$ grid. It is noted that the transfer function is symmetric for negative and positive $P_{in}$'s, and so we fold all the values into positive ones to increase our signal-to-noise.

The solid black diagonal line in Figure \ref{fig:sigloss_transfercurve} shows unity-transfer, or a perfect mapping from $P_{out}$ to $P_{in}$ (which we expect to occur for the unweighted case). The solid grey horizontal lines mark the peak of the data distribution $\hat{\textbf{p}}$ (the distribution is created from the $20$ bootstraps) for both the inverse covariance weighted and unweighted cases. One can eyeball that a data value of $10^{5} \, mK^{2} (h^{-1} Mpc)^{3}$, for example, would map approximately to a value of $\sim10^{8} \, mK^{2} (h^{-1} Mpc)^{3}$ using the transfer curve for the inverse covariance weighted case, implying a signal loss factor of $\sim1000$.

\begin{figure*}
	\centering
	\includegraphics[width=1\textwidth]{plots/sigloss_transfercurve.png}
	\caption{Signal loss transfer functions showing the relationship of $P_{in}$ and $P_{out}$, as defined by Equations \eqref{eq:Pin} and \eqref{eq:sigloss}. Smoothed out distributions (via kernel density estimators) are shown as colored heat-maps for the cases of inverse covariance weighted PAPER-64 data (left) and unweighted data (right). The solid black diagonal line marks a perfect unity mapping, and the solid grey horizontal line denotes the peak of $\hat{\textbf{p}}$, the data distribution. From these plots, it is clear that inverse covariance weighting results in $\sim3$ orders of magnitude of signal loss whereas the unweighted case does not exhibit loss.}
	\label{fig:sigloss_transfercurve}
\end{figure*}

@@

The first component is merely the fact that the values of the cross-terms in Equation \eqref{eq:pout_expand} change for different EoR signals, $\textbf{e}$. Since we draw different random $\textbf{e}$'s for every bootstrap, there is a (small) range of $P_{out}$ values that could result for the same $P_{in}$. This randomness is what causes the spread seen in the unweighted case (right plot of Figure \ref{fig:sigloss_transfercurve}), as well as most of the spread in the weighted case (left plot).

The second component of scatter is less intuitive and only applies (at a low level) to the weighted case. The origin of this scatter stems from how much the random EoR signal $\textbf{e}$ happens to look like the data $\textbf{x}$. For example, if we are extremely unlucky, $\textbf{e}$ would have the same shapes, or eigenvectors, as $\textbf{x}$. If weighting using empirically-derived covariances, this entire EoR signal would be destroyed. On the contrary, the less that $\textbf{e}$ looks like $\textbf{x}$, the less signal loss that would result. This range of how `unlucky' we can get is not that dramatic in real life, but it is important to correct for the fact that a particular $P_{out}$ value could in principle stem from a range of $P_{in}$ values.

@@

Beginning with smoothed probability distributions of the transfer function for the two cases (weighted and unweighted), we fit a polynomial to each, representing mean signal loss functions. We then account for the second component of scatter as described previously by de-convolving out $P_{out, unweighted}$ from $P_{out, weighted}$ for every $P_{in}$ value. In other words, we only account for the extra scatter that the weighted case has over the unweighted case, which, as can be seen in Figure \ref{fig:sigloss_transfercurve}, is not large.

Equipped with our new signal loss curves (a one-to-one mapping function for the unweighted case, and a somewhat smeared out function for the weighted case), we multiply our data distribution $\hat{\textbf{p}}$ element-wise to all $P_{in}$ columns, treating positive and negative halves of the data distribution separately. We note that for PAPER-64, $\hat{\textbf{p}}$ is sometimes narrowly peaked at positive power spectrum values (signifying detections of power), and sometimes centered around 0, representing noise-like measurements.

@@

To summarize this section, we also show power spectrum results for fringe-rate filtered PAPER-64 data before and after signal loss correction in Figure \ref{fig:ps2_data}, using inverse covariance weighting. The blue dashed line represents the unweighted power spectrum, which is identical in both panels (an important check, as we expect no signal loss for this case). The black and grey points are positive and negative power spectrum values plotted with $2\sigma$ error bars. Prior to signal loss correction, it is clear that the power spectrum is unfeasible because it is well below the theoretical noise level prediction (solid green curve). Post-correction, the power spectrum values blow up to be higher than both the theory and unweighted power spectrum.

Investigating this further, the reason why the post-corrected power spectrum is so high is due to the small number of independent samples in PAPER-64. PAPER-64 has $\sim 8$ independent modes after fringe-rate filtering while our power spectrum bandwidth has $21$ channels. Consequently, the eigenvalues of $\hat{\textbf{C}}$ cannot possibly describe real fluctuations beyond the first $8$. Hence, there is a substantial down-weighting of many random modes, a large loss that the signal loss transfer function captures and corrects for. \cc{elaborate?} In the next section, we investigate alternate weighting schemes to inverse covariance weighting, with the goal of finding one that balances the aggressiveness of down-weighting contaminants with minimizing the loss of EoR.

@@

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{plots/data_errors.png}
	\caption{$2\sigma$ power spectrum errors (not upper limits) using PAPER-64 data and three different bootstrapping methods. The data is fringe-rate filtered and a weighting matrix of $\textbf{I}$ (unweighted) is used in order to disentangle the effects of bootstrapping from signal loss. Three different bootstrapping methods are shown: the blue and black use methods where bootstrapping occurs on the baseline axis only (not the time axis), and they differ by how the baselines are randomly sampled. It is evident that sensitivity is maximized when baselines are sampled in a way that ensures mostly independent samples (only the last baseline spot is sampled randomly), as shown by the black curve. In red, bootstrapping occurs over both the baseline and time axes, illustrating how errors can be under-estimated if sampling more values than independent ones (fringe-rate filtering reduces the number of independent samples).}
	\label{fig:data_errors}
\end{figure}

@@

@@

\section{Acknowledgements}
The author would like acknowledge the UC Berkeley Chancellor's Fellowship, National Science Foundation Graduate Research Fellowship (Division of Graduate Education award 1106400), and thank Eric Switzer for helpful discussions. PAPER is supported by grants from the National Science Foundation (awards 0804508, 1129258, and 1125558). ARP, JCP, and DCJ would also like to acknowledge NSF support (awards 1352519, 1302774, and 1401708, respectively). JEA would like to acknowledge a generous grant from the Mount Cuba Astronomical Association for computing resources. We also graciously thank SKA-SA for site infrastructure and observing support.
\label{sec:Ack}

>>>>

To capture the full statistical likelihood of signal loss, one requires a quick way to generate many realizations of simulated $21$ cm signal visibilities. Here we use the same method used in \citet{ali_et_al2015}, where mock gaussian noise visibilities are filtered in fringe-rate space to retain only ``sky-like" time-modes. This signal is then added to the data.  \footnote{One specific change from \citet{ali_et_al2015} is that we add this simulated signal into the chain before the final fringe-rate filter is applied to the data. Previously, the addition was done after that final fringe-rate filter step.  This change results in an increased estimate of signal loss, likely due to the use of the fringe-rate filter as a simulator. Whether the loss would be estimated at the same level if different $21$ cm simulations were used remains an open question.}

@@

We are interested in quantifying how much variance in $\textbf{e}$ is lost after weighting $\textbf{r}$ and estimating the power spectrum according to OQE formalism. We investigate this by comparing two quantities we define as the input power spectrum and output power spectrum: $P_{in}$ and $P_{out}$, defined as (ignoring normalization factors):

@@

\begin{eqnarray}
\label{eq:sigloss}
P_{out,\alpha} &=& \hat{\textbf{p}}_{r,\alpha}-\hat{\textbf{p}}_{x,\alpha} \nonumber \\
&\propto& \textbf{r}^{\dagger}\textbf{R}_{r}\textbf{Q}^{\alpha}\textbf{R}_{r}\textbf{r} - \textbf{x}^{\dagger}\textbf{R}_{x}\textbf{Q}^{\alpha}\textbf{R}_{x}\textbf{x}.
\end{eqnarray}

$P_{in}$ represents the unweighted power spectrum of only $\textbf{e}$, our simulated EoR signal. $P_{out}$ is the weighted power spectrum of $\textbf{e}$ that would result from our pipeline if the signal was mixed with our data. Comparing the two quantities yields insight into how much of $\textbf{e}$ is lost due to our choice of weighting.

@@

One may wonder why $P_{out}$ cannot be computed simply as the weighted power spectrum of $\textbf{e}$ alone, namely $P_{out,\alpha} \propto \textbf{e}^{\dagger}\textbf{R}_{r}\textbf{Q}^{\alpha}\textbf{R}_{r}\textbf{e}$. In short, this naive expression fails to include correlations between the EoR signal and data, which emerges as statistical noise that is added to the estimate, decreasing the value of $P_{out}$. Mathematically, we can show this by expanding out Equation \eqref{eq:sigloss}:

@@

We see that $P_{out}$ is comprised of multiple terms. Taking the case of very large $\textbf{e}$ so that any terms involving only $\textbf{x}$ are small, yields:

@@

We see that our naive expression for $P_{out}$ is the first term in Equation \eqref{eq:pout_expand}, but there are also two additional terms. An initial assumption would be that the cross-terms that involve both $\textbf{e}$ and $\textbf{x}$ should be zero, since foregrounds and the cosmological signal are statistically unrelated. In \citet{ali_et_al2015}, this assumption is used, and the first term in Equation \eqref{eq:pout_expand} is directly compared to the input (unweighted) power spectrum to compute signal loss.

However, deeper investigations of these terms reveal that they contain non-negligible power. In fact, foreground modes and signal modes are anti-correlated on average. This negative cross-term power arises from inverting a covariance matrix with a strong diagonal and small off-diagonal terms, as we expect to be the form for $\textbf{C}_{r}$. Through the inversion process, the off-diagonal terms become negative and ... This effect is derived mathematically and explained more in depth in \citet{switzer_et_al2015}.

The spurious correlation between $\textbf{e}$ and $\textbf{x}$ means that we cannot compute signal loss using a signal-only simulation, which would yield greater values for $P_{out}$ and thereby underestimate signal loss. Therefore, in our revised signal loss computation we use the full quantity for $P_{out}$ as defined in Equation \eqref{eq:sigloss}, which subtracts the weighted power spectrum of the data from the weighted power spectrum of data plus EoR.

The relationship between the input and output power spectra, $P_{in}$ and $P_{out}$, can be thought of as a transfer function which, for a sampling of $P_{in}$ and $P_{out}$ provides a mapping from an input power spectrum distribution into an output distribution. By viewing data through this signal loss lens, we may then ask the question ``what input power spectrum distribution could this (signal-loss affected) data come from?" In the next section, we describe the shape of our signal loss transfer function and detail the computation used to translate our power spectrum result into one viewed through a signal loss lens.

@@

We now shift our attention towards computing signal loss for the fringe-rate filtered PAPER-64 dataset. While our methodology outlined below is independent of weighting scheme, here we demonstrate the computation using inverse covariance weighting ($\textbf{R} = \textbf{C}^{-1}$), the weighting scheme used in \citet{ali_et_al2015} which leads to substantial loss. With this weighting, our expressions for $P_{in}$ and $P_{out}$ become:

@@

Our signal loss transfer function for one $k$ value is shown in Figure \ref{fig:sigloss_transfercurve} and compares the weighted case ($\textbf{R} = \textbf{C}^{-1}$, left plot) with an unweighted case ($\textbf{R} = \textbf{I}$, right plot). The figures depict a kernel density estimation of our input to output power spectra. To sample the distribution, we run $40$ total injection levels, where the amplitude of \textbf{e} is increased in even increments in log-space and ranges from $P_{in} \sim -10^{13} \, mK^{2} (h^{-1} Mpc)^{3}$ (negative $P_{in}$ is injected when one copy of \textbf{e} in Equation \eqref{eq:Pin} is negative and the other is positive) to $P_{in} \sim10^{13} \, mK^{2} (h^{-1} Mpc)^{3}$. For each injection, we perform $20$ bootstraps over baselines, as explained in Section \ref{sec:Boot}, resulting in a total of $800$ data points on our $P_{in}$ vs. $P_{out}$ grid. We have verified that the transfer function is symmetric for negative and positive $P_{in}$'s, and so we fold all the values into positive ones to increase our signal-to-noise.

The solid black diagonal line in Figure \ref{fig:sigloss_transfercurve} shows unity-transfer, which we expect to occur for the unweighted case ($\textbf{R} = \textbf{I}$). The solid grey horizontal lines mark the peak of the data distribution $\hat{\textbf{p}}$ (the distribution is created from the $20$ bootstraps) for both the inverse covariance weighted and unweighted cases. One can eyeball that a data value of $10^{5} \, mK^{2} (h^{-1} Mpc)^{3}$, for example, would map approximately to a value of $\sim10^{8} \, mK^{2} (h^{-1} Mpc)^{3}$ using the transfer curve for the inverse covariance weighted case, implying a signal loss factor of $\sim1000$.

\begin{figure*}
	\centering
	\includegraphics[width=1\textwidth]{plots/sigloss_transfercurve.png}
	\caption{Signal loss transfer functions showing the relationship of $P_{in}$ and $P_{out}$, as defined by Equations \eqref{eq:Pin} and \eqref{eq:sigloss}. Kernel density estimations of the transfer functions are shown as colored heat-maps for the cases of inverse covariance weighted PAPER-64 data (left) and unweighted data (right). The solid black diagonal line marks a perfect unity mapping, and the solid grey horizontal line denotes the peak of $\hat{\textbf{p}}$, the data distribution. From these plots, it is clear that inverse covariance weighting results in $\sim3$ orders of magnitude of signal loss for power spectrum values above $\sim$$2 \times 10^{4}$ $mK^{2} \, (h^{-1} \, Mpc)^{3}$, whereas the unweighted case does not exhibit loss.}
	\label{fig:sigloss_transfercurve}
\end{figure*}

@@

The first component arises from the fact that the values of the cross-terms in Equation \eqref{eq:pout_expand} change for different mock EoR signals, $\textbf{e}$. Since we draw different random $\textbf{e}$'s for every bootstrap, there is a range of $P_{out}$ values that results from the same $P_{in}$. This randomness causes the spread seen in the unweighted case (right plot of Figure \ref{fig:sigloss_transfercurve}), as well as most of the spread in the weighted case (left plot).

The second component of scatter is less intuitive and only applies (at a low level) to the weighted case, but represents intrinsic scatter in signal loss. The origin of this scatter stems from how much the random EoR signal $\textbf{e}$ happens to look like the data $\textbf{x}$. For example, if we are extremely unlucky, one realization of $\textbf{e}$ would have the same shapes, or eigenvectors, as $\textbf{x}$. If weighting using empirically-derived covariances, this entire EoR signal would be destroyed. On the other hand, the less that $\textbf{e}$ looks like $\textbf{x}$, the less signal loss that would result. The intrinsic scatter we can get is not a dominant factor in this case but it is important to correct for the fact that a particular $P_{out}$ value could arise from a range of $P_{in}$ values.

@@

Beginning with smoothed probability distributions of the transfer function for the two cases (weighted and unweighted), we fit a polynomial to each, representing mean signal loss functions. We then account for the second component of scatter as described previously by de-convolving out $P_{out, unweighted}$ from $P_{out, weighted}$ for each $P_{in}$ value. In other words, we only account for the extra scatter that the weighted case has over the unweighted case, which, as can be seen in Figure \ref{fig:sigloss_transfercurve}, is not large.

Equipped with our new signal loss curves (a one-to-one mapping function for the unweighted case, and a somewhat smeared out function for the weighted case), we multiply our data distribution $\hat{\textbf{p}}$ element-wise to all $P_{in}$ columns, treating positive and negative halves of the data distribution separately. We note that for PAPER-64, $\hat{\textbf{p}}$ is sometimes narrowly peaked at positive power spectrum values (signifying detections of power), and sometimes centered around zero, representing measurements that are consistent with zero.

@@

To conclude this section, we also show power spectrum results for fringe-rate filtered PAPER-64 data before and after signal loss correction in Figure \ref{fig:ps2_data}, using inverse covariance weighting. The blue dashed line represents the unweighted power spectrum, which is identical in both panels (an important check, as we expect no signal loss for this case). The black and grey points are positive and negative power spectrum values plotted with $2\sigma$ error bars. Prior to signal loss correction, it is clear that the power spectrum is unfeasible because it is well below the theoretical noise level prediction (solid green curve). Post-correction, the power spectrum values blow up to be higher than both the theory and unweighted power spectrum.

Investigating this further, the reason why the post-corrected power spectrum is so high is due to the small number of independent samples in PAPER-64. PAPER-64 has eight independent modes after fringe-rate filtering while our power spectrum bandwidth has $21$ channels. Consequently, the eigenvalues of $\hat{\textbf{C}}$ cannot possibly describe real fluctuations beyond the first $8$. Hence, there is a substantial down-weighting of many random modes, a large loss that the signal loss transfer function captures and corrects for. \cc{elaborate?} In the next section, we investigate alternate weighting schemes to inverse covariance weighting, with the goal of finding one that balances the aggressiveness of down-weighting contaminants with minimizing the loss of EoR.

@@

\begin{figure}
	\centering
	\includegraphics[trim={0.3cm 0.3cm 0.3cm 0.3cm},width=\columnwidth]{plots/data_errors.png}
	\caption{$2\sigma$ power spectrum errors (not upper limits) using PAPER-64 data and three different bootstrapping methods. The data is fringe-rate filtered and a weighting matrix of $\textbf{I}$ (unweighted) is used in order to disentangle the effects of bootstrapping from signal loss. Three different bootstrapping methods are shown: the blue and black use methods where bootstrapping occurs on the baseline axis only (not the time axis), and they differ by how the baselines are randomly sampled. It is evident that sensitivity is maximized when baselines are sampled in a way that ensures mostly independent samples (only the last baseline spot is sampled randomly), as shown by the black curve. In red, bootstrapping occurs over both the baseline and time axes, illustrating how errors can be under-estimated if sampling more values than independent ones (fringe-rate filtering reduces the number of independent samples).}
	\label{fig:data_errors}
\end{figure}

@@

\cc{Should we even include this next section? I included it to follow-up on Section 2.3.1, but it's not saying anything new...}

@@

\section{Acknowledgements}
The author would like acknowledge the UC Berkeley Chancellor's Fellowship, National Science Foundation Graduate Research Fellowship (Division of Graduate Education award 1106400), and thank Eric Switzer for helpful discussions. PAPER is supported by grants from the National Science Foundation (awards 0804508, 1129258, and 1125558). ARP, JCP, DCJ, and JEA would also like to acknowledge NSF support (awards 1352519, 1302774, 1401708, and 1455151, respectively). We also graciously thank SKA-SA for site infrastructure and observing support.
\label{sec:Ack}
